---
title: "ggplot homework"
author: "Taavi Päll"
date: "3/19/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

```{r}
mpg
```


3. Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?


# Color
```{r}
ggplot(mpg) +
  geom_point(mapping = aes(x = cty, y = hwy, color = displ))
```

# Size
```{r}
ggplot(mpg) +
  geom_point(mapping = aes(x = cty, y = hwy, size = displ))
```

# Shape
```{r}
ggplot(mpg) +
  geom_point(mapping = aes(x = cty, y = hwy, shape = displ))
```


4. What happens if you map the same variable to multiple aesthetics?

```{r}
ggplot(mpg) +
  geom_point(mapping = aes(x = cty, y = hwy, color = displ, size = displ))
```


5. What does the stroke aesthetic do? What shapes does it work with? (Hint: use ?geom_point)

- For shapes that have a border (like 21),

```{r}
?geom_point
ggplot(mpg) +
  geom_point(mapping = aes(x = cty, y = hwy, stroke = displ, color = cyl), shape = 21)
```


6. What happens if you map an aesthetic to something other than a variable name, like aes(colour = displ < 5)?

```{r}
ggplot(mpg) +
  geom_point(mapping = aes(x = cty, y = hwy, color = displ < 5))
```

