---
title: "Data visualisation using ggplot2"
subtitle: "R stats TAY"
author: "Taavi Päll"
date: "2018-09-25"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(message = FALSE, fig.align = 'center')
```


class: inverse, center, middle

# Intro

---

## Why data visualization matters

- Visualization of your analysis results or data must be effective and present your results and data honestly

- Numerical summaries, like arithmetic mean or correlation coefficient, can be deceiving when shape of the data is not known or shown

???

- think of average salary or **datasaurus**

---

## Start every analysis by visualization

- Data visualization is natural starting point of every data analysis workflow

- Let's have look at the **datasaurus** dataset
    - here we have x and y variables for 13 variables (away, bullseye, etc.) 

```{r}
library(datasauRus)
datasaurus_dozen_wide
```

---

## Statistical properties of each dataset

- What is the correlation between x and y in different datasets?
- Here we calculate correlation coefficient for each dataset:
```{r}
library(dplyr)
datasaurus_dozen %>% 
  group_by(dataset) %>% 
  summarise(`Pearson's r` = cor(x, y))
```

---

## How are x and y related to each other?

- We can see that all these datasets display similar relationship between x and y variable
- What about mean and standard deviation?
```{r}
datasaurus_dozen %>% 
  group_by(dataset) %>% 
  summarise_at(vars(x, y), funs(mean, sd))
```

---

## Conclusion???

- We have dozen datasets with similar statistical properties
- We can conclude that all these datasets have same parameters and come from from the same data generating process
- As a final nuisance, let's create a some kind of plot (you know, reviewer...)

---

```{r}
library(ggplot2)
ggplot(datasaurus_dozen) +
  geom_point(ggplot2::aes(x, y)) +
  facet_wrap(~ dataset)
```

---
class: middle, inverse

> "never trust summary statistics alone; always visualize your data", since, while the data exhibits normal seeming statistics, plotting the data reveals a picture of a dinosaur. [Same Stats, Different Graphs: Generating Datasets with Varied Appearance and Identical Statistics through Simulated Annealing](https://www.autodeskresearch.com/publications/samestats)


