---
title: "Data transformation using dplyr (aka five verbs)"
author: "Taavi Päll"
date: "2018-03-11"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

I our previous classes we were working with small and cleaned up **mgp** dataset to go through steps of creating some of the most common visualisation types. 

In your data analysis workflow you need to plot out data at two points:

1. During **exploratory data analysis** where you learn to know your dataset and 
2. **reporting** when you try to communicate what you have found. 

However, data analysis is not a linear process, but an iterative process, where you go in circles number of times after you have done some "tweaking" (transformation and/or modeling) of your data. 

Here we add another stop to data analysis process: data transformation.

In most cases you need to transform your data during analysis, because in real life you rarely start with a dataset that is in the right form for visualisation and modeling. 


Usually you will need to:

- summarise your data, 
- create new variables, 
- rename variables, 
- reorder the observations. 


We are going to use the dplyr library from tidyverse to learn how to carry out these tasks. 

## Sources

Again, we are going to follow closely R4DS book chapter "Data transformation" available from http://r4ds.had.co.nz/transform.html. More examples are available from https://rstats-tartu.github.io/lectures/tidyverse.html#dplyr-ja-selle-viis-verbi


Load tidyverse library and dataset:

```{r}
library(tidyverse)
library(lubridate)
```

### Viruses data 

Viruses data contains virus strain taxonomic and genome info, dataset was downloaded from NCBI. Please see README.md in data folder.

Let's import viruses.csv file from data subfolder: 
```{r}
(viruses <- read_csv("data/viruses.csv"))
```


## dplyr basics

Most of the data transformation tasks can be carried out using five verbs from dplyr library:

- Pick observations by their values (filter()).
- Reorder the rows (arrange()).
- Pick variables by their names (select()).
- Create new variables with functions of existing variables (mutate()).
- Collapse many values down to a single summary (summarise()).

- These can all be used in conjunction with group_by() which changes the scope of each function from operating on the entire dataset to operating on it group-by-group. 


These six functions provide the verbs for a language of data manipulation.

All verbs work similarly:

- The first argument is a data frame.

- The subsequent arguments describe what to do with the data frame, using the variable names (without quotes).

- The result is a new data frame.

Together these properties make it easy to chain together multiple simple steps to achieve a complex result. Let's dive in and see how these verbs work.


## Filter rows with filter()

filter() allows you to subset observations based on their values. The first argument is the name of the data frame. The second and subsequent arguments are the expressions that filter the data frame. 

For example, we can filter bacterial viruses:

```{r}

```

dplyr runs the filtering operation and returns a new data frame. 

dplyr functions never modify their inputs, so if you want to save the result, you'll need to use the assignment operator, <-, like so:

Assign bacterial viruses to object phages:
```{r}

```

### Comparisons

What is this == operator? Why not use = to check equality:
```{r, eval=FALSE}
filter(viruses, host = "bacteria")
```

It appears that = is another assignment operator besides ->

There's another common problem you might encounter when using ==: floating point numbers. 

Although, theoretically TRUE, following comparisons return FALSE!
```{r}
sqrt(2) ^ 2 == 2
1/49 * 49 == 1
```

This is because computers and R use finite precision arithmetic and cannot store an infinite number of digits.

This can be overcome by using near() function instead of ==:
```{r}
near(sqrt(2) ^ 2,  2)
near(1 / 49 * 49, 1)
```

### Logical operators

Multiple comparisons within filter() function are combined with comma "," which means "and" (&). In case of "and" all comparisons must evaluate to TRUE for observations to be returned.

Together, logical (boolean) operators are:

- & is AND, 
- | is OR, 
- ! is NOT

The following code finds all viruses with genome info release date in 2019 OR in 2018, note that we use year function from lubridate package to extract year of release from date:
```{r}

```

You can't write something like filter(flights, month == 11 | 12) and in case of numeric years this will give you wrong answer instead of Error, so be careful:
```{r, eval=FALSE}
filter(viruses, year(release_date) == 2019 | 2018)
```

A useful short-hand for this problem is x %in% y. This will select every row where x is one of the values in y:

Filter observations where year of release_date is 2019 or 2018:
```{r}

```


Remember that (De Morgan's law):   

- !(x & y) is the same as !x | !y  
- !(x | y) is the same as !x & !y  


For example, if you wanted to find viruses with genome sizes between 1 to 2 kb, you could use either of the following two filters, note that we are excluding viruses with genomes smaller than 1kb or bigger than 2kb:
```{r}
filter(viruses, !(size_kb < 1 | size_kb > 2))
filter(viruses, !size_kb < 1, !size_kb > 2)
```

### Missing values

One important feature of R that can make comparison tricky are missing values, or NAs ("not availables"). 

NA represents an unknown value so missing values are "contagious": almost any operation involving an unknown value will also be unknown.

```{r}
NA > 5
10 == NA
NA + 10
NA / 2
```

As Rsudio already might suggest, if you want to determine if a value is missing, use is.na():
```{r}
x <- NA
is.na(x)
```

Let's use is.na() within filter to filter rows with missing GC content data (variable called "gc"):
```{r}

```

Ok. Now we got all rows with missing gc... how would you change this code to really exclude these rows with missing data (Hint: !FALSE):
```{r}

```


There are other functions that remove rows with NA in any of the columns from your data frame and keep only "complete cases":
```{r}
na.exclude(viruses)
na.omit(viruses)
drop_na(viruses)
```


## Arrange rows with arrange()

arrange() works similarly to filter() except that instead of selecting rows, it changes their order.

It takes a data frame and a set of column names to order by. If you provide more than one column name, each additional column will be used to break ties in the values of preceding columns:
```{r}
arrange(viruses, size_kb)
```

Use desc() to re-order by a column in descending order:
```{r}

```

Missing values are always sorted at the end, even with desc() function:
```{r}
df <- data_frame(x = c(5, NA, 2))
arrange(df, x)
```

```{r}
arrange(df, desc(x))
```

### Exercises

1. How could you use arrange() to sort all missing values to the start? (Hint: use is.na()).

```{r}
arrange(transactions, !is.na(price_total))
```


2. Sort transactions to find the busiest months (most transactions, arv) in each county (maakond).

```{r}

```

We'll come to this question back later...

3. Sort transactions to find the months with highest mean price payed (`pinnaühiku hind(eur /m2) keskmine`) for each conty.

```{r}

```

come back again...


4. Which type of apartements (`pindala(m2)`) sold the best in 2017? Which sold worst in 2017?

```{r}

```


## Select columns with select()

select() allows you to rapidly zoom in on a useful subset of columns using operations based on the names of the variables.

Select first three columns:
```{r}
select(transactions, 1:3)
```

Select columns from year to count:
```{r}
select(transactions, year:count)
```

!!! select works in both directions: L>R and R<L!
select year, county, area and count
```{r}
select(transactions, year, county, area, count)
```


Exlude columns year, county, area and count:
```{r}
select(transactions, -year, -county, -area, -count)
```

```{r}
select(transactions, -c("year", "county", "area", "count"))
```

Use minus sign to exclude variables! You can submit variables to select as character vector.


There are a number of __helper functions you can use within select()__:

- starts_with("abc"): matches names that begin with "abc".

Select columns that start with "price"
```{r}
select(transactions, starts_with("price"))
```


- ends_with("xyz"): matches names that end with "xyz".


Select columns that end with "max"
```{r}
select(transactions, ends_with("max"))
```


- contains("ijk"): matches names that contain "ijk".

All columns that contain word "price"
```{r}
select(transactions, contains("price"))
```


- matches("(.)\\1"): selects variables that match a regular expression. This one matches any variables that contain repeated characters. You'll learn more about regular expressions in strings.

```{r, eval = FALSE}
matches("^abc") # same as starts_with("abc)
matches("xyz$") # same as ends_with("abc)
matches("ijk") # same as contains("ijk)
```


```{r}
select(transactions, matches("[[:punct:]]"))
```


- num_range("V", 1:3) matches V1, V2 and V3.

- everything() is useful if you have a handful of variables you'd like to move to the start of the data frame.


Move columns year and month to the start of the data frame, and keep all other columns.
```{r}
select(transactions, year, month, everything())
```


See ?select for more details.


### Exercises

1. What happens if you include the name of a variable (e.g. count) multiple times in a select() call?

```{r}
select(transactions, count, count, count)
```


2. What does the one_of() function do? Why might it be helpful in conjunction with this vector?
```{r}
(vars <- c("area", "count", "fffffff"))
select(transactions, one_of(vars))
```

```{r, eval=FALSE}
select(transactions, c("area", "count", "ffffff"))
```


3. Select from 'transactions' all variables that contain string 'PRICE' (note case!). 

```{r}
select(transactions, contains("PRICE"))
```


Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default?
```{r}
?select
trans_price_case <- select(transactions, contains("PRICE", ignore.case = FALSE))
class(trans_price_case)
```

