---
title: "Data transformation using dplyr (aka five verbs)"
author: "Taavi Päll"
date: "2018-03-11"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

I our previous classes we were working with small and cleaned up **mgp** dataset to go through steps of creating some of the most common visualisation types. 

In data analysis workflow you need to plot out data at two points:

1. During **exploratory data analysis** where you learn to know your dataset and 
2. **reporting** when you try to communicate what you have found. 

Importantly, ata analysis is not a linear process, but an iterative process of data transformation, modeling and visualisation. 



Here we add another step to data analysis process: data transformation.

In most cases you need to transform your data during analysis, because in real life you rarely start with a dataset that is in the right form for visualisation and modeling. 


Usually you will need to:

- summarise your data, 
- create new variables, 
- rename variables, 
- reorder the observations. 


We are going to use the dplyr library from tidyverse to learn how to carry out these tasks. 

## Sources

Again, we are going to follow closely R4DS book chapter "Data transformation" available 

- from http://r4ds.had.co.nz/transform.html.    
- More examples are available from https://rstats-tartu.github.io/lectures/tidyverse.html#dplyr-ja-selle-viis-verbi


Load tidyverse library and dataset:

```{r}
library(tidyverse)
library(lubridate) # library to work with dates and time
library(here) # (always) locate files in your project folder P.S. load here after lubridate, because lubridate has also (now deprecated) function called here
```

### Viruses data 

Viruses data contains virus strain taxonomic and genome info, dataset was downloaded from NCBI. Please see README.md in data folder.


```{r}
download.file("https://raw.githubusercontent.com/rstats-tartu/datasets/master/viruses.csv", here("data", "viruses.csv"))
```

Let's import viruses.csv file from data subfolder: 
```{r}
(viruses <- read_csv(here("data", "viruses.csv")))
```


here() uses clever heuristics to identify your working directory and updates file paths respectively:

```{r}
here("data", "viruses.csv")
```

Set up here like so, cd to your working directory and run:
```{r}
here::set_here()
```


You could also just run:
```{r, eval=FALSE}
viruses <- read_csv("data/viruses.csv")
```


## dplyr basics

Most of the data transformation tasks can be carried out using five verbs from dplyr library:

- Pick observations by their values (filter()).
- Reorder the rows (arrange()).
- Pick variables by their names (select()).
- Create new variables with functions of existing variables (mutate()).
- Collapse many values down to a single summary (summarise()).

- These can all be used in conjunction with group_by() which changes the scope of each function from operating on the entire dataset to operating on it group-by-group. 


These six functions provide the verbs for a language of data manipulation.

All verbs work similarly:

- The first argument is a data frame.

- The subsequent arguments describe what to do with the data frame, using the variable names (without quotes).

- The result is a new data frame.

Together these properties make it easy to chain together multiple simple steps to achieve a complex result. Let's dive in and see how these verbs work.


## Filter rows with filter()

filter() allows you to subset observations based on their values.

The first argument is the name of the data frame. The second and subsequent arguments are the expressions that filter the data frame. 

For example, we can filter bacterial viruses:

```{r}
filter(viruses, host == "bacteria")
```

dplyr runs the filtering operation and returns a new data frame. 

dplyr functions never modify their inputs, so if you want to save the result, you'll need to use the assignment operator, <-, like so:

Assign bacterial viruses to object phages:
```{r}
phages <- filter(viruses, host == "bacteria")
phages
```

### Comparisons

What is this == operator? Why not use = to check equality:

```{r, eval=FALSE}
filter(viruses, host = "bacteria")
```

It appears that = is another assignment operator besides ->

There's another common problem you might encounter when using ==: floating point numbers. 

Although, theoretically TRUE, following comparisons return FALSE!

```{r}
sqrt(2) ^ 2 == 2
1/49 * 49 == 1
```

This is because computers and R use finite precision arithmetic and cannot store an infinite number of digits.

This can be overcome by using near() function instead of ==:
```{r}
near(sqrt(2) ^ 2,  2)
near(1 / 49 * 49, 1)
```

### Logical operators

Multiple comparisons within filter() function are combined with comma "," which means "and" (&). 

In case of "and" all comparisons must evaluate to TRUE for observations to be returned.

Together, logical (boolean) operators are:

- & is AND, 
- | is OR, 
- ! is NOT



The following code finds all viruses with genome info release date in 2019 OR in 2018, 
note that we use year function from lubridate package to extract year of release from date:

```{r}
year(ymd("2017 08 27")) # lubridate library

filter(viruses, year(release_date) == 2019 | year(release_date) == 2018)
```

You can't write something like filter(flights, month == 11 | 12) and in case of numeric 
years this will give you wrong answer instead of Error, so be careful:

```{r, eval=FALSE}
filter(viruses, year(release_date) == 2019 | 2018)
```

A useful short-hand for this problem is x %in% y. This will select every row where x is one of the values in y:

Filter observations where year of release_date is 2019 or 2018:
```{r}
filter(viruses, year(release_date) %in% c(2019, 2018))
```

What happens under the hood:
```{r, eval=FALSE}
year(viruses$release_date) %in% c(2019, 2018)
```


Remember that (De Morgan's law):   

- !(x & y) is the same as !x | !y  
- !(x | y) is the same as !x & !y  


For example, if you wanted to find viruses with genome sizes between 1 to 2 kb, you could use either of the following two filters, note that in the first case we are excluding viruses with genomes smaller than 1kb or bigger than 2kb:

```{r}
filter(viruses, !(size_kb < 1 | size_kb > 2))
```

Previous expression can be rewritten like so:
```{r}
filter(viruses, !(size_kb < 1), !(size_kb > 2))
```


```{r}
filter(viruses, size_kb >= 1, size_kb <= 2)
```

Which one from these three is more explicit? Note that comma here in filter function between logical evaluations means "&".

"Small than", "bigger than" an "NOT":
```{r}
3 >= c(2, 3, 4)
3 <= c(2, 3, 4)
3 != c(2, 3, 4)
```




### Missing values

One important feature of R that can make comparison tricky are missing values, or NAs ("not availables"). 

NA represents an unknown value so missing values are "contagious": 

almost any operation involving an unknown value will also be unknown.

```{r}
NA > 5
10 == NA
NA + 10
NA / 2
```

As Rsudio already might suggest, if you want to determine if a value is missing, use is.na():
```{r}
x <- NA
is.na(x)
```

Let's use is.na() within filter to filter rows with missing GC content data (variable called "gc"):
```{r}
filter(viruses, is.na(gc))
```

Ok. Now we got all rows with missing gc... how would you change this code to really exclude these rows with missing data (Hint: !FALSE):
```{r}
filter(viruses, !is.na(gc))
```


There are other functions that remove rows with NA in any of the columns from your data frame and keep only "complete cases":

```{r}
na.exclude(viruses)
na.omit(viruses)
drop_na(viruses)
```

### Finding non-exact matches

Often you find yourself in need to filter categorical variables based on some non-exact matching, for example when 

- values are too long,   
- there are too many unique unkown values,   
- some observations that belong to the same category have slightly different values (`foo bar` and `foo-bar` and `foobar`). 

To solve this problem more elegantly, you can use regular expressions. (for help see ?`regular expression`)

regular expressions are covered in *tidyverse* by the **stringr** package. 


The stringr function that is useful within filter is str_detect(). Although, stringr belongs to tidyverse, it is not loaded together with core tidyverse packages by default, so you need to load it separately.


```{r}
library(stringr)
fruits <- c("banana", "foo TRUMP bar", "foo bar", "foo-bar", "foobar", "foo")
str_detect(fruits, "foo(.*)?bar")
```

As you can see, str_detect() returns logical vector.

So how this should work when you run str_detect inside filter?

Salmonalla viruses:
```{r}
filter(viruses, str_detect(organism_name, "Salmonella"))
```



### Exercises

1. Find all viruses that:

- have genome of over 10000kb

```{r}

```

- belong to *Papillomaviridae*

```{r}

```

- are from plants and environment

```{r}

```

- were released between 01. January 1980 - 01. January 1990, including these days:

```{r}
filter(viruses, release_date > "1980-01-01", release_date < "1990-01-01")
```

2. there is also between() function in dplyr. What does it do? How can you use it to find viruses with genomes between 1 to 2 kb?
```{r}

```

3. How many viruses belong to "unclassified" subgroup:

```{r}

```


```{r}
filter(viruses, str_detect(str_to_lower(organism_name), "enterovi"))
```


## Arrange rows with arrange()

arrange() works similarly to filter() except that instead of selecting rows, it changes their order.

It takes a data frame and a set of column names to order by. 

If you provide more than one column name, each additional column will be used to break ties in the values of preceding columns:

```{r}
arrange(viruses, size_kb)
```

Use desc() to re-order by a column in descending order. You can combine variables:
```{r}
arrange(viruses, organism_name, desc(size_kb))
```

Missing values are always sorted at the end, even with desc() function:
```{r}
df <- tibble(x = c(5, NA, 2))
arrange(df, x)
```

```{r}
arrange(df, desc(x))
```


### Get unique rows with distinct()

Sometimes observations become duplicated during data wrangling, sometimes you need to get unique combinations of observations. 
dplyr has distinct() function to retain only unique rows from input table.


viruses data has `r nrow(viruses)` rows, but what if we would like to get only unique organism_name and tax_id combinations...  
First let's select organism_name and tax_id columns
```{r}
taxids <- select(viruses, 1:2)
taxids
```

Here we are... with distinct organism_name tax_id combinations.
```{r}
distinct(taxids)
```

distinct() works on tables
```{r}
?distinct
```

unique() function from base R works on vectors
```{r}
x <- rep(letters[1:3], each = 3) # simulate character vector x with repetitions
x
unique(x) # select unique set of values from x
```


### Exercises

1. How could you use arrange() to sort viruses with missing gc values to the start? (Hint: use is.na()).

```{r}
arrange(viruses, !is.na(gc))
```


2. Sort viruses to find viruses with highest GC% content:

```{r}
arrange(viruses, desc(gc))
```


## Select columns with select()

select() allows you to rapidly zoom in on a useful subset of columns using operations based on the names of the variables.

Select first three columns:
```{r}
select(viruses, 1:3)
```

Select columns from group to host:
```{r}
select(viruses, group:host)
select(viruses, host:group)
```

!!! select works in both directions: L>R and R<L!



select organism_name, size_kb, gc:
```{r}
select(viruses, organism_name, size_kb, gc)
```


Exlude columns bioproject_accession, bioproject_id, segments:
```{r}
select(viruses, -bioproject_accession, -bioproject_id, -segments)
```

Another way, exclude as a vector, in case of vector use quoted variable names!:
```{r}
vars_out <- c("bioproject_accession", "bioproject_id", "segments")
select(viruses, -vars_out)
```

This expression will fail, because we don't have objects called bioproject_accession etc in our R environment!
```{r, eval=FALSE}
vars_out <- c(bioproject_accession, bioproject_id, segments)
```


Use minus sign to exclude variables! 

> Submit variables to select as character vector!


There are a number of __helper functions you can use within select()__:

- starts_with("abc"): matches names that begin with "abc".

Select columns that start with "bio"
```{r}
select(viruses, starts_with("bio"))
```


- ends_with("xyz"): matches names that end with "xyz".


Select columns that end with "date"
```{r}
select(viruses, ends_with("date"))
```


- contains("ijk"): matches names that contain "ijk".

All columns that contain word "project"
```{r}
select(viruses, contains("project"))
```


- matches("(.)\\1"): selects variables that match a regular expression. 

This one matches any variables that contain repeated characters. You'll learn more about regular expressions in strings.

```{r, eval = FALSE}
matches("^abc") # same as starts_with("abc")
matches("xyz$") # same as ends_with("xyz")
matches("ijk") # same as contains("ijk")
```

Select columns/variables with punctuation marks: 
```{r}
select(viruses, matches("[[:punct:]]"))
```


- num_range("V", 1:10000) matches V1, V2 and V3.


- everything() is useful if you have a handful of variables you'd like to move to the start of the data frame.

Move columns tax_id, size_kb and gc to the start of the data frame, and keep all other columns.

You can rearrange the order of columns.
```{r}
select(viruses, tax_id, size_kb, gc, everything())
```


See ?select for more details.


### Exercises

1. What happens if you include the name of a variable (e.g. gc) multiple times in a select() call?

```{r}
select(viruses, gc, gc, gc)
```


2. What does the one_of() function do? 

Why might it be helpful in conjunction with this vector?
```{r}
(vars <- c("tax_id", "gc", "fffffff"))
select(viruses, one_of(vars))
```

What happens if you try to select columns just by using this vector:

```{r, eval=FALSE}
select(viruses, c("tax_id", "gc", "fffffff"))
```

3. Select all variables from 'viruses' dataset that contain string 'GC' (note case!). 

Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default?

```{r}
?select
select(viruses, contains("GC", ignore.case = TRUE))
```

