---
title: "Data transformation using dplyr (aka five verbs) 1"
author: "Taavi Päll"
date: "09. oktoober 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
I our previous classes we have been working with small cleaned up dataset to go through steps of creating some of the most common visualisation types. 

In your workflow you are going to need data visualisation at two points:

(1) during exploratory data analysis where you learn to know your dataset and 
(2) during report preparation when you try to communicate what have you found. 

This is not two stop trip, it's more like a roundabout, an iterative process, where you pass these two point multiple times after you have done some "tweaking" of your data. By "tweaking" I mean here data transformation and/or modeling. 

You need to transform your data during analysis, because in real life you rarely start with a dataset that is in the right form for visualisation and modeling. So, often you will need to:

- summarise your data or to 
- create new variables, 
- rename variables, or 
- reorder the observations. 

We are going to use the dplyr library from tidyverse to learn how to carry out these tasks. 

## Sources

Again, we are going to follow closely R4DS book chapter "Data transformation" available from http://r4ds.had.co.nz/transform.html. More examples are available from https://rstats-tartu.github.io/lectures/tidyverse.html#dplyr-ja-selle-viis-verbi

## Class V

Load tidyverse library and dataset:
```{r}

```

### Estonian apartment transactions data. 
Transactions data contain monthly apartment sales data from January 2005 to October 2018 split up by counties and size of apartments. Price info is available when at least five transactions has been carried out.

Let's import transactions_with_consumer_price_index.csv file from data subfolder: 
```{r}

```


## dplyr basics

Most of the data transformation tasks can be carried out using five verbs from dplyr library:

- Pick observations by their values (filter()).
- Reorder the rows (arrange()).
- Pick variables by their names (select()).
- Create new variables with functions of existing variables (mutate()).
- Collapse many values down to a single summary (summarise()).

- These can all be used in conjunction with group_by() which changes the scope of each function from operating on the entire dataset to operating on it group-by-group. 


These six functions provide the verbs for a language of data manipulation.

All verbs work similarly:

- The first argument is a data frame.

- The subsequent arguments describe what to do with the data frame, using the variable names (without quotes).

- The result is a new data frame.

Together these properties make it easy to chain together multiple simple steps to achieve a complex result. Let’s dive in and see how these verbs work.

## Prepare data, cont. 

Let's improve these terrible column names:
```{r}

```


## Filter rows with filter()

filter() allows you to subset observations based on their values. The first argument is the name of the data frame. The second and subsequent arguments are the expressions that filter the data frame. 

For example, we can select data from 'Tartu maakond':
```{r}

```

dplyr runs the filtering operation and returns a new data frame. 

dplyr functions never modify their inputs, so if you want to save the result, you'll need to use the assignment operator, <-:

Assign data from 'Tartu maakond' to object tartu_mk:
```{r}

```

### Comparisons

What is this == operator? Why not use = to check equality:
```{r, eval=FALSE}
filter(transactions, year = 2005)
```

It appears that = is another assignment operator besides ->

There's another common problem you might encounter when using ==: floating point numbers. Although, theoretically TRUE, following comparisons return FALSE!
```{r}
sqrt(2) ^ 2 == 2
1/49 * 49 == 1
```

This is because computers and R use finite precision arithmetic and cannot store an infinite number of digits.

This can be overcome by using near() function instead of ==:
```{r}
near(sqrt(2) ^ 2,  2)
near(1 / 49 * 49, 1)
```

### Logical operators

Multiple comparisons within filter() function are combined with comma "," which means "and" (&). In case of "and" all comparisons must evaluate to TRUE for observations to be returned.

Together, logical (boolean) operators are:

- & is AND, 
- | is OR, 
- ! is NOT

The following code finds all transactions in November OR December:
```{r}

```

You can’t write filter(flights, month == 11 | 12) and in case of numeric months this will give you wrong answer instead of Error, so be careful:
```{r, eval=FALSE}
filter(transactions, kuu == 11 | 12)
```

A useful short-hand for this problem is x %in% y. This will select every row where x is one of the values in y:

Filter observations where kuu (month) is Nov or Dec (we have numeric months in data frame):
```{r}

```

Sometimes you can simplify complicated subsetting by remembering De Morgan's law: !(x & y) is the same as !x | !y, and !(x | y) is the same as !x & !y. 
For example, if you wanted to find months with 5 to 10 transactions, you could use either of the following two filters:
```{r}
filter(transactions, !(arv < 5 | arv > 10))
filter(transactions,  !arv < 5, !arv > 10)
```

### Missing values

One important feature of R that can make comparison tricky are missing values, or NAs ("not availables"). NA represents an unknown value so missing values are "contagious": almost any operation involving an unknown value will also be unknown.
```{r}
NA > 5
10 == NA
NA + 10
NA / 2
```

As Rsudio already might suggest, if you want to determine if a value is missing, use is.na():
```{r}
x <- NA
is.na(x)
```

Let's use is.na() within filter to remove rows with missing price info (`tehingu summa (eur) kokku`):
```{r}

```

Ok. Now we got all rows with missing price_total... how would you change this code to really exclude these rows with missing data (Hint: !FALSE):
```{r}

```

There is another function that works with data frames to find rows with complete set of observations - complete.cases():
```{r}

```

Note that you need to put dot inside complete.cases() function. Why so? 
You can try to feed transactions data directly to complete.cases(). How this output looks like?


### Exercises - homework

1. Find all transactions that

- Had an area_mean more than one hundred square meters
- Took place in Saare maakond
- Were done during summer (July, August, and September). Another useful dplyr filtering helper is between(). What does it do?

```{r}

```


2. How many rows have a missing total_price? What other variables are missing?

```{r}

```


3. Why is NA ^ 0 not missing? Why is NA | TRUE not missing? Why is FALSE & NA not missing? Can you figure out the general rule? (NA * 0 is a tricky counterexample!)
```{r}
NA ^ 0
NA | TRUE
```

```{r}
NA * 0
NA * FALSE
```


## Arrange rows with arrange()

arrange() works similarly to filter() except that instead of selecting rows, it changes their order.

It takes a data frame and a set of column names to order by. If you provide more than one column name, each additional column will be used to break ties in the values of preceding columns:
```{r}
arrange(transactions, arv)
```

Use desc() to re-order by a column in descending order:
```{r}
arrange(transactions, desc(arv))
```

Missing values are always sorted at the end, even with desc() function:
```{r}
df <- tibble(x = c(5, 2, NA))
arrange(df, x)
```

```{r}
arrange(df, desc(x))
```

### Exercises

1. How could you use arrange() to sort all missing values to the start? (Hint: use is.na()).

```{r}

```


2. Sort transactions to find the busiest months (most transactions, arv) in each county (maakond).

```{r}

```

We'll come to this question back later...

3. Sort transactions to find the months with highest mean price payed (`pinnaühiku hind(eur /m2) keskmine`) for each conty.

```{r}

```

come back again...


4. Which type of apartements (`pindala(m2)`) sold the best in 2017? Which sold worst in 2017?

```{r}

```


## Select columns with select()

select() allows you to rapidly zoom in on a useful subset of columns using operations based on the names of the variables.

Select three columns:
```{r}

```

Select columns from aasta to arv:
```{r}

```

```{r}

```


Exlude columns from `pindala(m2)` to :
```{r}

```

There are a number of __helper functions you can use within select()__:

- starts_with("abc"): matches names that begin with "abc".

- ends_with("xyz"): matches names that end with "xyz".

- contains("ijk"): matches names that contain "ijk".

- matches("(.)\\1"): selects variables that match a regular expression. This one matches any variables that contain repeated characters. You’ll learn more about regular expressions in strings.

```{r, eval = FALSE}
matches("^abc") # same as starts_with("abc)
matches("xyz$") # same as ends_with("abc)
matches("ijk") # same as contains("ijk)
```


- num_range("V", 1:3) matches V1, V2 and V3.

- everything() is useful if you have a handful of variables you'd like to move to the start of the data frame.

See ?select for more details.

Move columns aasta and kuu to the start of the data frame, keep all other columns.
```{r}

```

### Exercises

1. What happens if you include the name of a variable (e.g. arv) multiple times in a select() call?

```{r}

```


2. What does the one_of() function do? Why might it be helpful in conjunction with this vector?
```{r}
(vars <- c())
select(transactions, one_of(vars))
```

3. Select from 'transactions' all variables that contain string 'PRICE' (note case!). 

```{r}

```


Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default?
```{r}
?select
select(transactions, contains("PRICE", ignore.case = FALSE))
```

